Continuations via Negative Information Flow

### 2012-02-13 dual.tex
new formatting for the semantics
dgms
more diagrams and notes

\usepackage{graphicx}
\usepackage{longtable}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{mdwlist}
\usepackage{txfonts}
\usepackage{xspace}
\usepackage{amstext}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{multicol}
\usepackage[nodayofweek]{datetime}
\usepackage{etex}
\usepackage[all, cmtip]{xy}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{multicol}

@@

@@

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! Gx = \Gamma^{\times}
%! G = \Gamma
%! [] = \Box
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |--> = \mapsto
%! |- = \vdash
%! ==> = \Longrightarrow
%! <== = \Longleftarrow
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! ::= = &::=&
%! /= = \neq
%! forall = \forall
%! exists = \exists
%! empty = \epsilon
%! langRev = \Pi
%! theseus = Theseus

@@

\begin{abstract}
TODO
\end{abstract}

@@

\paragraph*{Int Construction.}

@@

@@

%subcode{bnf} include main
% Values, v = () | (v, v) | L v | R v
% Combinators, c &=& iso | c (;) c | c(+)c | c (*) c 
%
% Sequential Contexts, P, F = [] | P:c
% Parallel Contexts, D = [] 
%                     &|& (P, [](+)c, F):D 
%                     &|&  (P, c(+)[], F):D
%                     &|& (P, c v, F):D 
%                     &|&  (P, v c, F):D
%
%
% Machine States  = <P, F, c v, D> | <P, F, v c, D>
% Start State  = <[], [], c v, []>
% Stop State  = <P, [], v c, []>

Combinator reconstruction, {{P[c]}}: 
%subcode{opsem} include main
% [][c] '= c
% (P:c')[c] '= P[c'(;)c]

@@

\begin{itemize}
\item 
Basic reduction of an isomorphism. Note that the evaluation leaves the
adjoint of the combinator behind. This will become important when we
reverse the direction of computation.
%subcode{opsem} include main
% <P; F; iso v; D>      &|-->& <P; F; v'~ iso{dagger}; D>

\item
Sequencing involves pushing and popping from the Future and Past
continuations:
%subcode{opsem} include main
% <P, F, c1(;)c2 v , D>  &|-->& <P, F:c2, c1 v, D>
% <P, F:c2, v c1, D> &|-->& <P:c1, F, c2 v, D>

\item
Parallel composition, captures the current Future and Past and extends
a Dump/meta-continuation.
%subcode{opsem} include main
% <P, F, c1 (+) c2 (L v), D> &|-->& <[], [], c1 v, (P,[](+)c2,F):D>
% <P, F, c1 (+) c2 (R v), D> &|-->& <[], [], c2 v, (P,c1(+)[],F):D> 
% <P', [], v c1, (P,[](+)c2,F):D> &|-->& < P, F, (L v) (P'[c1](+)c2{dagger}), D > 
% <P', [], v c2, (P,c1(+)[],F):D> &|-->& <P, F, (R v) (c1{dagger}(+)P'[c2]), D>

\item
Similarly for products:
%subcode{opsem} include main
% <P, F, c1 (*) c2 (v1, v2), D> &|-->& <[], [], c1 v1, (P, [] (*) c2 v2,F):D>
% <P', [], v1 c1, (P,[] (*) c2 v2,F):D> &|-->& <[], [], c2 v2, (P,v1 P'[c1] (*) [],F):D>
% <P', [], v2 c2, (P,v1 c1 (*) [],F):D> &|-->& <P, F, (v1, v2) (c1(*)P'[c2]):D>
and symmetrical rules for evaluation along the second branch. 
%subcode{opsem} include main
% <P', [], v2 c2, (P,c1 v1 (*) [],F):D> &|-->& <[], [], c1 v1, (P,[] (*) v2 P'[c2],F):D>
% <P', [], v1 c1, (P,[] (*) v2 c2,F):D> &|-->& <P, F, (v1, v2) (P'[c1] (*) c2):D>

@@

@@

\item
Operational Semantics.
%subcode{opsem} include main
% <P; F; eps (R v); D>      &|-->& <F; P; (L (-v))~ eps; D{dagger}>

@@

\item
The adjoint of a Dump is defined to be:
%subcode{opsem} include main
% []{dagger} '= []
% ((P, [](+)c, F):D){dagger} '= (F, [](+)c{dagger}, P):D{dagger}
% ((P, c(+)[], F):D){dagger} '= (F, c{dagger}(+)[], P):D{dagger}
% ((P, [](*)c v, F):D){dagger} '= (F, [](*)v c, P):D{dagger}
% ((P, [](*)v c, F):D){dagger} '= (F, [](*)c v, P):D{dagger}
% ((P, c v(*)[], F):D){dagger} '= (F, v c(*)[], P):D{dagger}
% ((P, v c(*)[], F):D){dagger} '= (F, c v(*)[], P):D{dagger}

>>>>

\usepackage{graphicx}
\usepackage{longtable}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{mdwlist}
\usepackage{txfonts}
\usepackage{xspace}
\usepackage{amstext}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{multicol}
\usepackage[nodayofweek]{datetime}
\usepackage{etex}
\usepackage[all, cmtip]{xy}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{multicol}
\usepackage{bm}

@@

\newcommand{\lcal}{\ensuremath{\lambda}-calculus}
\newcommand{\G}{\ensuremath{\mathcal{G}}\xspace}

@@

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! Gx = \Gamma^{\times}
%! G = \Gamma
%! [] = \Box
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |--> = \mapsto
%! |- = \vdash
%! ==> = \Longrightarrow
%! <== = \Longleftarrow
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! -o = \multimap
%! ::= = &::=&
%! /= = \neq
%! forall = \forall
%! exists = \exists
%! empty = \epsilon
%! langRev = \Pi
%! langRevT = \Pi^{o}
%! langRevEE = \Pi^{\eta\epsilon}
%! theseus = Theseus
%! * = \times

@@

\begin{abstract}
We show a deep symmetry between functions and delimited continuations,
values and continuations that arises in {{langRev}} in a manner that
is reminiscent of Filinksi's Symmetric \lcal
~\cite{Filinski:1989:DCI:648332.755574}. The symmetry arises by
extending {{langRev}} with a notion of additive duality over the
monoid {{(+, 0)}} by including {{eta}} and {{eps}} operators of
Compact Closed Categories. The resulting dual types, which we denote
{{-b}}, have a time traveling ``backward information flow''
interpretation and allow for the encoding of higher-order function and
iteration via the construction of {{trace}} operators, thereby making
the extended language {{langRevEE}} a Turing-complete reversible
programming language with higher-order functions and first-class
delimited continuations.
\end{abstract}

@@

\paragraph*{Int Construction.}
For a traced monoidal category {{C}} the Int construction produces a
Compact Closed Category called Int {{C}} \cite{joyal1996traced}.
Further we know that the target of the Int construction is isomorphic
to the target of \G construction of Abramsky \cite{Abramsky96:0} from
Haghverdi. However, note that the {{langRevEE}} is not the same as the
image of the Int construction on {{langRevT}}, since the the later
lacks a multiplicative tensor that distrbutes over the additive tensor
in Int {{langRevT}}.

@@

\paragraph*{GoI machines} 
We can now encode the GoI machine of Mackie \cite{Mackie2011,DBLP:conf/popl/Mackie95}.

@@

%subcode{bnf} include main
% Values, v = () | (v, v) | L v | R v
% Combinators, c &=& iso | c (;) c | c(+)c | c (*) c 
%
% Sequential Contexts, P, F = [] | P:c
% Parallel Contexts, D = [] 
%                     &|& D:(P| [](+)c| F)
%                     &|& D:(P| c(+)[]| F)
%                     &|& D:(P| [] (*) c v| F)
%                     &|& D:(P| [] (*) v c| F)
%                     &|& D:(P| c v (*)  []| F)
%                     &|& D:(P| v c (*) []| F)
%
%
% Machine States  = D[P| c v| F] | D[P| v c| F]
% Start State  = [][[]| v c| []]
% Stop State  = [][P| c v| []]

Combinator reconstruction, {{P[c]}}:
%subcode{opsem} include main
% [](c) '= c
% P:c'(c) '= P(c'(;)c)

@@

\begin{itemize}
\item 
Basic reduction of an isomorphism. Note that the evaluation leaves the
adjoint of the combinator behind. This will become important when we
reverse the direction of computation.
%% %subcode{opsem} include main
%% % <P; F; iso v; D>      &|-->& <P; F; v'~ iso{dagger}; D>

%subcode{opsem} include main
% D[P|v iso| F]      &|-->& D[P| iso{dagger} v| F]

\item
Sequencing involves pushing and popping from the Future and Past
continuations:
%subcode{opsem} include main
% D[P|v (c1(;)c2)| F]  &|-->& D[P| v c1| F:c2]
% D[P| c1 v| F:c2] &|-->& D[P:c1| v c2| F]

\item
Parallel composition, captures the current Future and Past and extends
the parallel context.
%subcode{opsem} include main
% D[P|(L v) c1 (+) c2| F] &|-->& D:(P|[](+)c2|F)[ []| v c1| [] ]
% D[P| (R v) c1 (+) c2| F] &|-->& D:(P|c1(+)[]|F)[ []| v c2| [] ]
% D:(P|[](+)c2|F)[P'|c1 v| [] ]  &|-->& D[P|(P'(c1)(+)c2{dagger}) (L v)| F] 
% D:(P|c1(+)[]|F)[P'| c2 v| [] ] &|-->& D[P|(c1{dagger}(+)P'(c2)) (R v)| F]

\item
Similarly for products:
%subcode{opsem} include main
% D[P| (v1, v2) c1 (*) c2| F] &|-->& D:(P| [] (*) v2 c2|F)[ []| v1 c1| [] ]
% D:(P|[] (*) v2 c2|F)[P'| c1 v1| [] ] &|-->& D:(P|P'(c1) v1 (*) []|F)[ []| v2 c2| [] ]
% D:(P|c1 v1(*) []|F)[P'|c2 v2| [] ]  &|-->& D[P|c1(*)P'(c2) (v1, v2)| F]

and symmetrical rules for evaluation along the second branch. 
%subcode{opsem} include main
% D:(P|v1 c1 (*) []|F)[P'| c2 v2| [] ] &|-->& D:(P,[] (*) P'(c2) v2,F)[ []| v1 c1| [] ]
% D:(P|[] (*) c2 v2|F)[P'|c1 v1| [] ] &|-->& D[P |P'(c1) (*) c2 (v1, v2)| F]

@@

Note that we deviate from the categorical definition of {{eta}} and
{{eps}} slightly in that they swap the order of {{-b}} and {{b}} in
choice of {{eta}}. This however does not affect us, because we deal
with a symmetric category.

@@

\item
Operational Semantics.
%subcode{opsem} include main
% D[P| (R v) eps| F]      &|-->&  D{dagger}[F | eps (L (-v))| P]

@@

\item
The adjoint of a parallel context is defined to be:
%subcode{opsem} include main
% []{dagger} '= []
% (D:(P, [](+)c, F)){dagger} '= D{dagger}:(F, [](+)c{dagger}, P)
% (D:(P, c(+)[], F)){dagger} '= D{dagger}:(F, c{dagger}(+)[], P)
% (D:(P, [](*)c v, F)){dagger} '= D{dagger}:(F, [](*)v c, P)
% (D:(P, [](*)v c, F)){dagger} '= D{dagger}:(F, [](*)c v, P)
% (D:(P, c v(*)[], F)){dagger} '= D{dagger}:(F, v c(*)[], P)
% (D:(P, v c(*)[], F)){dagger} '= D{dagger}:(F, c v(*)[], P)

\section{Diagrams}

{{eta}}

\begin{center}
  \includegraphics{diagrams/eta.pdf}
\end{center}

Note that the connective is a {{+}}, hence only one of the branches
may be inhabited at any time. Thus the action of {{eta}} is to
transfer a backward flowing value on one wire to a forward flowing
value on the other wire.

{{eps}}

\begin{center}
  \includegraphics{diagrams/eps.pdf}
\end{center}

Coherence condition

\begin{center}
  \includegraphics{diagrams/coherence.pdf}
\end{center}

Function

\begin{center}
  \includegraphics{diagrams/function.pdf}
\end{center}

Let us use the shorthand {{b1 -o b2 = -b1 + b2}}

Function application

\begin{center}
  \includegraphics{diagrams/apply1.pdf}
\end{center}

\begin{center}
  \includegraphics{diagrams/apply2.pdf}
\end{center}

Function composition

\begin{center}
  \includegraphics{diagrams/compose1.pdf}
\end{center}

\begin{center}
  \includegraphics{diagrams/compose.pdf}
\end{center}

Delimited continuation

\begin{center}
  \includegraphics{diagrams/delimc.pdf}
\end{center}

\paragraph*{Observability.} 
Execution of program is defined by {{c : b1 <-> b2}} when evaluated on
input {{v1 : b1}} gives us a value {{v2 : b2}} on
termination. Execution is well defined only if {{b1}} and {{b2}} are
entirely positive types. Consider the program that

\begin{center}
  \includegraphics{diagrams/shuffle.pdf}
\end{center}

We define observables to be only positive types. The ouputs of
programs that ouput mixed positive and negative types are not
observable.  Also, programs that input mixed positive and negatives
types are not executable.

\subsection{To think about}

\begin{itemize}

\item {{(-b1) * b2}} is not the same as {{b1 * (-b2)}} as there is no
  isomorphism to translate from one to the other.

\item {{-(b1 * b2)}} is not the same as {{b1 * b2}} and is also not
  the same as {{(-b1) * (-b2)}} as there is no isomorphism that goes
  from one to the other.

\item {{-(-b)}} is not the same as {{b}} as there is no isomorphism
  that goes from one to the other.
